---
title: "R Notebook"
output: html_notebook
---

```{r}
rm(list=ls()) # remove the objects from environment
print(getwd()) # get working directory
print(ls()) # list created objects
n <- 1000 # create object n
x <- seq(1,n) # create sequence x
sum(x) # sum of all elements of sequence
```


In R, all values are stored as vectors.

```{r}
x <- 1:20 # create vector of 1 to 20

print(x)
```

## Data Types

R has five basic or atomic classes of objects. The `c()~ function can be usd to create vector of objects.

```{r}
x <- c(0.5, 0.6) # numeric vector
x <- c(TRUE, FALSE) # logical vector
x <- c(T, F) # logical
x <- c("a", "b", "c") # character vector
x <- c(1 + 0i, 2 + 4i) # complex
x <- vector("numeric", length=10) # vector of length 10 with values 0
y <- c(1.7, "a") # create character vector with coersion
y <- c(TRUE, 2) # numeric
y <- c("a", TRUE) # character vector
```

We can explicitly coerce objects using `as.*` functions.

```{r}
x <- 0.6
class(x) # numeric
as.numeric(x)
as.logical(x)
as.character(x)
x <- c("a", "b", "c")
print(as.numeric(x))
print(as.logical(x))
```


### Lists and Matrices

Lists are a special type of vector that can contains elements of different classes. Matrices are vectors with dimension attribute. Entries in columns are filled first, then second column and so on. This can be changed by `cbind` or `rbind` functions.

```{r}
x <- list(1, "a", TRUE, 1 + 4i)
m <- matrix(nrow=2, ncol=3)
print(dim(m))
print(attributes(m))
m <- matrix(1:6, nrow=2, ncol=3)
print(dim(m))
m <- 1:10
print(dim(m))
dim(m) <- c(2,5)
print(dim(m))

x <- 1:3
y <- 10:12
cbind(x,y)
rbind(x,y)
```

### Factors

Factor is a special type of vector to represent categorical data. Factors are treated specially by modelling functions like `lm()` and `glm()`. We can set order of levels using `levels` argument in `factor()` function.

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no")) # create factor
table(x) # represent in table format
unclass(x) # strips out class and represents as integers, usually in alphabetically level is assigned.
attr(x,"levels")
x <- factor(c("yes", "no", "yes", "yes", "no"), levels=c("yes", "no")) # here, yes will be baseline level
unclass(x)
```

### Missing values

Missing values are denoted by NA or NaN. NaN is also NA but reverse is not true.

```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
is.nan(x)
x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)
```

### DataFrame

DataFrame is used to store tabular data. They can store different classes objects in each column unlike objects. DataFrames also have special attributes `row.names` for annotating data row-wise. Data Frames are usually created by `read.table()` or `read.csv()`. They can be converted from matrix using `data.matrix()` however the result is coerced.

```{r}
x <- data.frame(foo=1:4, bar=c(T,T,F,F))
print(nrow(x))
print(ncol(x))
```

R objects can have names.

```{r}
x <- 1:3
names(x)
names(x) <- c("foo", "bar", "baz")
names(x)
length(x)

x <- list(a=1, b=3, c=4)
print(x)
m <- matrix(1:4, nrow=2, ncol=2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
print(m)
test <- c(4, TRUE)
class(test)
```


### Read csv file

```{r}
data <- read.csv("hw1_data.csv")
names(data)
head(data,2) # get first two records
nrow(data) # number of records
tail(data, 2) # check last 2 records
data[47,] # get 47th row
nrow(subset(data, is.na(Ozone)))
sub = subset(data, !is.na(Ozone), select=Ozone)
apply(sub, 2, mean)

sub <- subset(data, Ozone > 31 & Temp > 90, select=Solar.R)
apply(sub, 2, mean)

sub <- subset(data, Month == 6, select=Temp)
apply(sub, 2, mean)

sub = subset(data, Month == 5 & !is.na(Ozone), select=Ozone)
apply(sub, 2, max)
```

```{r}
install.packages("swirl") # install swirl package
packageVersion("swirl") # check version of the package
library("swirl") # load the library package for swirl
install_from_swirl("R Programming")
```

```{R}
swirl()
```
}